---
title: "RStudio Exercise 5"
author: "Tuomas Tiihonen"
date: "5 joulukuuta 2017"
output: html_document
---

# Dimensionality reduction

```{r setup chapter 5, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(corrplot)
library(magrittr)
library(FactoMineR)
library(tidyr)
library(ggplot2)
library(factoextra)
```
## Task 1 - read, explore, and describe
Let's first do some basic setup..
```{r Assignment 1, eval = TRUE, echo = TRUE, include = TRUE}
setwd('Z:/IODS-project/data/')
human <- read.csv("human.txt", header = TRUE)
rownames(human) <- human$X
human <- subset(human, select = -X)
```

Then we'll explore the dataset.
```{r Structure of human, echo = TRUE}
str(human)
```
The dataset consists of the following data for each of the countries:

"Edu2.FM" = Ratio of females to each male with at least secondary education  
"Labo.FM" = Ratio of females to each male in the labour force  
"Edu.Exp" = Expected years of schooling  
"Life.Exp" = Life expectancy at birth  
"GNI" = Gross National Income per capita in US Dollars  
"Mat.Mor" = Maternal mortality rate  
"Ado.Birth" = Adolescent birth rate  
"Parli.F" = Percetange of female representatives in parliament  


```{r Dimensions of human}
dim(human)
```
The dataset contanins 155 observations and 8 variables.


## Task 2 - Graphical overview and summaries
```{r Correlation plot}
cor_matrix<-cor(human) %>% round(digits=2)
cor_matrix
corrplot.mixed(cor_matrix, upper="square", tl.cex = 0.7)
```

From the correlation plot we can see that the percentage of female representatives in parliament and the ratio of females to males in the labour force do not correlate strongly with any of the other variables ($\mathit{r} \le \pm .25$). Their strongest correlation is with each other.

The strongest correlation coefficients ($\mathit{r} \ge\pm.7$) can be found between  

* Maternal mortality rate and life expectancy ($\mathit{r}=-.86$)  
* Expected years of schooling and life expectancy ($\mathit{r}=.79$)  
* Maternal mortality rate and adolescent birth rate ($\mathit{r}=.76$)  
* Expected years of schooling and maternal mortality rate ($\mathit{r}=-.74$)  
* Adolescent birth rate and life expectancy ($\mathit{r}=-.73$)  
* Expected years of schooling and adolescent birth rate ($\mathit{r}=-.7$)  

```{r Summary}
summary(human)
```

The distributions of the variables conform quite well with normal distribution. GNI and maternal mortality rate seem a bit suspect, as the former has a maximum well beyond the first three first quartiles, and the latter is skewed towards the third quartile.

## Task 3 - PCA 

```{r PCA with unstandardised data}
pca_human <- prcomp(human)
biplot(pca_human, choices = 1:2, cex = 0.4)
```

## Task 4 - Standardisation and new PCA

```{r Standardisation and new PCA}
human_std <- scale(human)
pca_human_std <- prcomp(human_std)
biplot(pca_human_std, choices = 1:2, cex = 0.4)
```

The unstandardised plot does not offer substantial information in addition to the large variance in GNI, and the GNI outliers that cause the unreadability of the plot. From the GNI we can deduce that oil-rich countries are leading the pack with Singapore and Luxembourg. The effect of the GNI dominance in the first principal component (PC1) makes the second and subsequence principal components undiscernible.  

Standardising the variables make the principal component graph more readable by diminishing the dominance of GNI.

From the standardised biplot we can find three dimensions of interconnected variables: GNI, life expectancy, expected years of schooling and the ratio of females to males with at least secondary education are grouped together; the ratio of females to males in the labour force are grouped together, and maternal mortality rate and adolescent birth rate are grouped together. These groupings correspond to the correlation coefficients analysed earlier.  

## Task 5 - Interpretations

When analysed through principal component 1, the first grouping is a polar opposite to the third grouping, with both adolescent births and maternal mortality decreasing when GNI, life expectancy, and education levels increase.  



## Task 6 - MCA of tea dataset
```{r Tea time data read}
tea <- read.table("http://factominer.free.fr/book/tea.csv",header=TRUE,sep=";")
```

```{r Tea time structure}
str(tea)
```

We can now see that all of the variables except age are factors with 2-7 levels.

```{r Tea time dimensions}
dim(tea)
```
There are 300 observations of 36 variables.

```{r Tea time column drop and plot, fig.width = 12, fig.height = 12}
tea <- subset(tea, select = -age)
gather(tea) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

The graphical representation of the variables shows, that there are some categories with very low frequencies that may distort the analysis. Hence the following variables should be dropped before engaging in further analysis:

* after.dinner
* after.lunch
* home
* iron.absorption
* slimming

```{r MCA 1}
tea <- subset(tea, select = -c(after.dinner, after.lunch, home, iron.absorption, slimming))
mca <- MCA(tea, graph = FALSE)
summary(mca)
fviz_screeplot(mca, addlabels = TRUE)
```

```{r MCA 2}
fviz_mca_var(mca, choice = "mca.cor", 
            repel = TRUE, # Avoid text overlapping (slow)
            ggtheme = theme_minimal())
fviz_mca_biplot(mca, repel = TRUE, ggtheme = theme_minimal())
# Contributions of rows to dimension 1
fviz_contrib(mca, choice = "var", axes = 1, top = 15)
# Contributions of rows to dimension 2
fviz_contrib(mca, choice = "var", axes = 2, top = 15)
#biplot(mca, choices = 1:2, cex = 0.6)
fviz_mca_var(mca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # avoid text overlapping (slow)
             ggtheme = theme_minimal()
             )
plot(mca, habillage = "quali", invisible=c("ind"))
```